---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RNentropy

<!-- badges: start -->
[![R-CMD-check](https://github.com/Federico77z/RNentropy/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Federico77z/RNentropy/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This is the implementation of a method based on information theory devised for the identification of genes showing a significant variation of expression across multiple conditions. Given expression estimates from any number of  RNA-Seq samples and conditions it identifies genes or transcripts with a  significant variation of expression across all the conditions studied, together with the samples in which they are over- or under-expressed. [Zambelli F. et al. (2018)](https://doi.org/10.1093/nar/gky055).

A detailed walk-through on how to use RNentropy is available at [Zambelli F., Pavesi G. (2021)](https://doi.org/10.1007/978-1-0716-1307-8_6).

The `main` branch is the development line based on the CRAN 1.2.3 release. The exact release sources are preserved by the [`v1.2.3`](https://github.com/Federico77z/RNentropy/tree/v1.2.3) tag.

## Installation

Install the stable release from CRAN:

``` r
install.packages("RNentropy")
```

Install the development version from GitHub:

``` r
pak::pak("Federico77z/RNentropy")
```

## Example

This is a basic example showing how to use RNentropy. Please see [Zambelli F., Pavesi G. (2021)](https://doi.org/10.1007/978-1-0716-1307-8_6) for more info.

```{r example}
library(RNentropy)
# basic example code
##load expression values and experiment design
data("RN_Brain_Example_tpm", "RN_Brain_Example_design")
#Run RNentropy 
Results <- RN_calc(RN_Brain_Example_tpm, RN_Brain_Example_design)
#select only genes with significant changes of expression
Results <- RN_select(Results)
#Compute the Point Mutual information Matrix
Results <- RN_pmi(Results)
```

## Isoform-switch analysis

The package includes a compact subset of the historical S7 data used to
validate the isoform-switch implementation. The analysis can be run directly
from the expression file with `RNentropy_iso_switch()`:

```{r isoswitch, eval=FALSE}
S7_File <- system.file("extdata", "isoswitch_s7.tsv", package = "RNentropy")
S7_Results <- RNentropy_iso_switch(
  S7_File,
  tr.col = "TR_ID",
  gene.col = "GENE_ID"
)
S7_Results$gene_status
S7_Results$sample_status
S7_Results$lpv
```

The same calculation can be run on a data frame with `RN_iso_calc()`:

```{r isoswitch-data-frame, eval=FALSE}
S7_Data <- read.delim(S7_File, check.names = FALSE)
S7_Calculated <- RN_iso_calc(S7_Data, gene.col = "GENE_ID")

identical(S7_Results$gene_status, S7_Calculated$gene_status)
all.equal(S7_Results$lpv, S7_Calculated$lpv)
```

## Independent C++ implementation

The independently developed C++ reference implementation, including the historical isoform-switch executable, is archived at [Federico77z/RNentropy_cpp](https://github.com/Federico77z/RNentropy_cpp).
